# PART II: Managing biological data {#lab2}

**Getting started**

- Review [best practices](https://ahurford.github.io/quant-guide-all-courses/style.html) for guidance on how to organize your files.

- Download, to an appropriate folder, the following .csv files from Brightspace:

    -   shrub-volume-data.csv

    -   shrub-volume-experiments.csv

    -   shrub-volume-sites.csv

    -   surveys.csv

    -   species.csv

    -   plots.csv

- Open and save a new R Script. Complete the information at the start of the R Script  [template](https://ahurford.github.io/quant-guide-all-courses/style.html#template).

- Set your [working directory](https://ahurford.github.io/quant-guide-all-courses/rintro.html#working-directory)

- Install and load the [dplyr](https://dplyr.tidyverse.org/) R package. 

If you're unsure how to install a package, see [R Packages](https://ahurford.github.io/quant-guide-all-courses/rintro.html#r-packages).

Throughout these instructions are links to the relevant sections in [*Quantitative Skills for Biology*](https://ahurford.github.io/quant-guide-all-courses/).

If you do not complete these steps, you will not be able to complete the **HAND IN** questions.

**TO HAND-IN**
You are to hand in a `.R` file, formatted as shown in [Best  Practices - Template](https://ahurford.github.io/quant-guide-all-courses/style.html#template) where the lines of code to produce the output asked for under **HAND-IN** are produced in the order they are asked for.

------------------------------------------------------------------------

## Grouping and Joining Data

### Shrub Volume Aggregation 

In this section, we revisit [grouping data](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#group-summary) and using [pipes](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#pipes). The following code calculates the mean width of plants at each site. Both the pipe operation method and the non-pipe operation method will be highlighted.

```{r, include=F}
library(dplyr)
```

```{r}
# attach dplyr
library(dplyr)

# load the data 
shrub_dims <- read.csv("lab_data/shrub-volume-data.csv")

# Calculate the mean width for each site:
## 1. Without using pipes:
by_site <- group_by(shrub_dims, site)
avg_width <- summarize(by_site, avg_width=mean(width))

## 2. Using pipes:
avg_width_pipe <- shrub_dims %>%
  group_by(site) %>%
  summarize(mean(width))

# Showing the first six rows with both methods
head(avg_width)
head(avg_width_pipe)
```

**HAND IN. [Q1]** Add a line of code to your R script where you use *pipes* to calculate the mean *height* of plants in each *experiment*, and use `head()` to print the output (the above code is intended as a guide you can modify).

**HAND IN. [Q2]** Add a line of code to your R script where you use *pipes* to calculate the maximum, `max()`, *height* of a plant at each *site* and use `head()` to print the output.

### Shrub Volume Join 

In addition to the main data table on shrub dimensions, there are two additional data tables:

- `shrub-to-volume-experiment.csv`, which describes the manipulation for each experiment; and - `shrub-volume-sites.csv`, which provides information about the different sites.

Your are to use `inner_join()` to create a new data frame with the site details added. You can read about `inner_join()` [here](https://stat545.com/join-cheatsheet.html).


**HAND IN. [Q3]**. In your R Script, indicate with a comment that you are starting a new section `## Shrub volume join`.

Add lines of code to your R Script where you:

- Import the experiments data, `shrub-volume-experiment.csv` and use `head()` to view the first six rows of data.

- View the first six rows of the data `shrub-volume-data.csv` (note that in the above code this was assigned the variable name `shrub_dims`).

**HAND IN. [Q4]** In your R Script, add a commented out line of code where you provide the name of columns that appear in both data sets.

**HAND IN. [Q5]** In your R Script, add a commented out line of code where you provide the *values* of the column with the same name that appear in both data sets.

**HAND IN. [Q6]** In your R Script, add a line of code where you use the `inner_join()` function to combine the sites data with the shrub dimensions, `shrub_dims` data (note you may have given a different variable name to your import of `shrub-volume-data.csv`). 

**HAND IN. [Q7]** In your R Script, add lines of code to import the sites data, `shrub-to-volume-sites`, and then combine these data with both the data on shrub dimensions and the experiment data to produce a **single data frame that contains all of the data.** Hand in the `tail()` output of the new data frame.  *Hint: you've just created a single data frame that contains both the shrub and experimental data.*

```{r, include=F, echo=F}
##2. Shrub Volume Join

#import the experiments data and combine it with the shrub dimensions data

experiments <- read.csv("lab_data//shrub-volume-experiments.csv")
shrub_combined <- inner_join(shrub_dims, experiments, by = "experiment")

#import the sites data and combine with both data on dimensions and experiment to produce a single data frame
sites <- read.csv("lab_data/shrub-volume-sites.csv")
shrub_full <- inner_join(shrub_combined, sites, by = "site")
```

## Portal Data Aggregation 

This exercise uses the `survey.csv` (Portal Teaching Database survey table) file you downloaded for last week's lab. If the `survey.csv` is not in your workspace download it from last week's Brightspace folder. 

In this section you will be using the `group_by()`, `summarize()`, `tally()` and `filter()` functions. More information about each of the functions can be found under the Help Tab or through the `?function` and `??function` commands. Examples for how to use the `?` and `??` command are below.

```{r}
?group_by
??group_by
```

*Hint*: For this section it maybe useful to use pipe `%>%` operations and the `tally` function. For example, I can tally the number of individuals that either identify as `M` or `F` in the sex column.

```{r}
survey %>% 
  group_by(sex) %>%
  tally()
```

**HAND IN** Do the following: 

1. Use the `group_by()` function to get a count of the number of individuals in each species `ID`. Include either the output or place into a table and include. 

2. Use the `group_by()` function to get a count of the number of individuals in each species ID in each year. *Hint: you can group_by two variables*. Include either the output or place into a table and include. 

3. Use the `filter()`, `group_by()`, and `summarize()` functions to get the mean mass of species `DM` in each year. Inlcude the `tail` output. *Hint: double check the `survey` data is ready to use.*

```{r, echo=F, include=F}
#import surveys data
surveys <- read.csv("lab_data/surveys.csv")

#get a count of the number of individuals in each species ID (##Add hint to help files about using "tally" function and makign use of pipes)
surveys %>%
  group_by(species_id) %>%
  tally()


#get a count of the number of individuals in each species ID in each year (##ADD hint about how to do grouping by two variables)
surveys %>%
  group_by(species_id, year) %>%
  tally()

#get mean mass of species DO in each year (TRICKY thing here is that they're not told to filter out the weights that are "NA")

surveys %>%
  filter(species_id == "DO") %>%
  filter(!is.na(weight)) %>%
  group_by(year) %>%
  summarize(avg_mass = mean(weight))
```

## Fix the code 

This section is a follow up to the Shrub Volume Aggregation.

The following code is supposed to import the shrub volume data and calculate the average shrub volume for each site, and separately, for each experiment.

```{r, error=T, eval=F, warning=FALSE}

read.csv(shrub-volume-data.csv)
shrub_data %>%
  mutate(volume = length * width * height) %>%
  group_by(site) %>%
  summarise(mean_volume= max(volume))

shrub_data %>%
  mutate(volume = length * width * height) %>%
  group_by(experiment) %>%
  sumarize(mean_volume = mean(volume))

# This is an example of a comment within code
```

**HAND IN**: 

1. Fix the errors in the code so that it does what it's supposed to. 

2. Add comments `#` throughout to indicate what you have corrected. A portion of your mark for this question qill be determined by the quality of the comments. 

## Portal Data Joins {#portdatajoins}

This exercise also uses the `survey.csv` you downloaded from last week's lab, in addition to the files listed below. Make sure that you have downloaded and loaded these data frames into your R. 

-   [species.csv](%3Chttps://ndownloader.figshare.com/files/3299483)

-   [plot.csv](https://ndownloader.figshare.com/files/3299474)

**HAND IN:** Using the data sets above, answering the following question

1.  Use `inner_join()` to create a table that contains the information for both the `survey` table and the `species` table. Include both the head and tail output of the new data frame. 

2.  Use `inner_join()` twice to create a table that contains the information from all three tables. Include both the head and the tail output of the new data frame. 

3.  Use `inner_join()` and `filter()` to get a data frame with the information from the `survey` and `plots` tables where the `plot_type` is `Control`. Include only the tail output of the new data frame.

```{r, include=F, echo=F}
surveys <- read.csv("lab_data/surveys.csv")
species <- read.csv("lab_data/species.csv")

#new table combining info in surveys and species data frames

species_surveys <- inner_join(surveys, species, by = "species_id")


#tables that combine info from species, surveys, plots

plots <- read.csv("lab_data/plots.csv")
species_survey_plots <- inner_join(species_surveys, plots, by = "plot_id")

#data from from surveys and plots tables where plot_type is "control"

plots %>%
  filter(plot_type == "Control") %>%
  inner_join(surveys, by = "plot_id")
```

## Portal Data `dplyr` Review 

The following questions will be using the same data frames as \@ref(portdatajoins) above:`surveys.csv`, `species.csv`, and `plot.csv`.

**HAND IN** We want you to do an analysis comparing the size of individuals on the `Control` plot to the `Long-term Krat Exclosure`. **Create a data frame** with the `year`, `genus`, `species`, `weight`, and `plot_type` for all cases where the plot type is either `Control` or `Long-term Krat Exclosure`. Only include cases where `Taxa` is `Rodent`. Remove any records where the `weight` is missing. Include the head output and the `length()` of the new data set. entire *Hint: it maybe easiest to use the pipe %>% operation to create*

```{r, include=F, echo=F}

newTable<-species %>%
  filter(taxa == "Rodent") %>%
  inner_join(surveys, species, by = "species_id") %>%
  filter(!is.na(weight)) %>%
  inner_join(plots, by = "plot_id") %>%
  filter(plot_type == "Control"| plot_type == "Long-term Krat Exclosure") %>%
  select(year, genus, species, weight, plot_type) %>%
  arrange(year)

```

## Extracting vectors from data frames 

Using the `survey.csv` (Portal Teaching Database survey table):

**HAND IN** Do the following

1) Use `$` to extract the weight column into a vector and rename into `weight_vector` 

2) Use [,] to extract the month column into a vector and rename into `month_vector`. *Recall that in [x,y] x corresponds to the row, and y corresponds to the column.* 

3) Extract the `hindfoot_length` column into a vector and calculate the mean `hindfoot_length` ignoring null values.

## Building data frames from vectors

You have data on the `length`, `width`, and `height` of 10 individuals of the English yew *Taxus baccata* stored in the following vectors:

```{r}
length <- c(2.2, 2.1, 2.7, 3.0, 3.1, 2.5, 1.9, 1.1, 3.5, 2.9)
width <- c(1.3, 2.2, 1.5, 4.5, 3.1, NA, 1.8, 0.5, 2.0, 2.7)
height <- c(9.6, 7.6, 2.2, 1.5, 4.0, 3.0, 4.5, 2.3, 7.5, 3.2)
```

**HAND IN** a data frame that contains these three vectors as columns along with a genus column containing the name *Taxus* for all rows **and** a species column containing the word *baccata* for all rows.

```{r, include=F, echo=F}
length <- c(2.2, 2.1, 2.7, 3.0, 3.1, 2.5, 1.9, 1.1, 3.5, 2.9)
width <- c(1.3, 2.2, 1.5, 4.5, 3.1, NA, 1.8, 0.5, 2.0, 2.7)
height <- c(9.6, 7.6, 2.2, 1.5, 4.0, 3.0, 4.5, 2.3, 7.5, 3.2)

NewDataFrame <- data.frame(length, width, height) %>%
  mutate(Genus = "Taxus") %>%
  mutate(species = "baccata") %>%
  select(Genus, species, length, width, height)
```
