---
title: "BIOL 1002: Principles of Biology 1002"
author: "Jake Prosser, Yolanda Wiersma, Amy Hurford"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook: default
  #bookdown::pdf_book: default
---
# Before coming to the labs

```{r, include=F, echo=F}
# additional information not include in on the website
library(dplyr)

```


**Overview:** In these two computer labs, we will be building on the R skills you learned in BIOL 1001. We will be focusing on manipulating and organizing data in tables, and learn a useful package called [dplyr](https://dplyr.tidyverse.org/) (pronounced "d-plier")


Install:

- [R and RStudio](https://ahurford.github.io/quant-guide-all-courses/install.html#installing) (if you have not already).

Read:

- [Introduction to R - Workflow](https://ahurford.github.io/quant-guide-all-courses/rintro.html#workflow)

- [Introduction to R - Working directory](https://ahurford.github.io/quant-guide-all-courses/rintro.html#working-directory)

-   [Entering and loading data - Loading or importing data](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#loading-or-importing-data)

-   [Entering and loading data - Inspecting the data](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#inspecting-the-data)

- [Best practices](https://ahurford.github.io/quant-guide-all-courses/style.html)

- [Handling data](https://ahurford.github.io/quant-guide-all-courses/handling-data.html)

**Remember!** The [*Quantitative Skills for Biology*](https://ahurford.github.io/quant-guide-all-courses/) manual you used in BIOL 1001 is still online (and updated regularly). If you need to review how to install and use R, please the manual! You can access it here: <https://ahurford.github.io/quant-guide-all-courses/>.

When using R, you have many ways to get [help](https://ahurford.github.io/quant-guide-all-courses/help.html).

In addition, these labs are based on _Week 3: Data in Tables_ (PART I) and _Week 4: Grouping & Joining Data_ (PART II) of the online course [Data Carpentry for Biologists](https://datacarpentry.org/semester-biology/schedule/) by Dr. Ethan White. If you are not sure how to execute a particular command, you can refer to the _Lecture_ material on [Dr. White's website](https://datacarpentry.org/semester-biology/lectures/R-data/)</u> and/or his [instructional videos](https://datacarpentry.org/semester-biology/lectures/R-data/) that go with the relevant lessons. The example data in the lecture material and videos is different from the data you use in the exercise.

**After these labs you should be able to:**

- install packages and load data,
-   understand the data manipulation functions of `dplyr`,
-   and execute a simple import and data filter scenario

-----------------------------------------------------------

# PART I: Managing biological data

**Getting started**

- Review [best practices](https://ahurford.github.io/quant-guide-all-courses/style.html) for guidance on how to organize your files.

- Download, to an appropriate folder, the following .csv files from Brightspace:

    -   shrub-volume-data.csv

    -   max-temperature.csv

    -   surveys.csv

- Open and save a new R Script. Complete the information at the start of the R Script  [template](https://ahurford.github.io/quant-guide-all-courses/style.html#template).

- Set your [working directory](https://ahurford.github.io/quant-guide-all-courses/rintro.html#working-directory)

- Install and load the [dplyr](https://dplyr.tidyverse.org/) R package. 

If you're unsure how to install a package, see [R Packages](https://ahurford.github.io/quant-guide-all-courses/rintro.html#r-packages).

Throughout these instructions are links to the relevant sections in [*Quantitative Skills for Biology*](https://ahurford.github.io/quant-guide-all-courses/).

If you do not complete these steps, you will not be able to complete the **HAND IN** questions.

**TO HAND-IN**
You are to hand in a `.R` file, formatted as shown in [Best  Practices - Template](https://ahurford.github.io/quant-guide-all-courses/style.html#template) where the lines of code to produce the output asked for under **HAND-IN** are produced in the order they are asked for.

-----------------------------------------------------------

## Shrub Volume Data Basic

After writing the command `library(dplyr)` in your R script to attach `dplyr`, next [import the data](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#loading-or-importing-data). If you use the programmatic way (i.e., command line), your command may look something like this:

```{r}
ShrubData <- read.csv("lab_data/shrub-volume-data.csv")
```

The exact command will depend on where you saved `shrub-volume-data-data.csv` and how you have set your [working directory](https://ahurford.github.io/quant-guide-all-courses/rintro.html#working-directory). You may call your dataset `ShrubData` or give it a different name.

You will need this command in your R script or you will not be able to complete the **HAND-IN** questions that appear next.

**HAND-IN. [Q1]**  Add a line of code to your R script where you use the [`select()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#selecting-a-column) function to select the `site` and `experiment` columns of `shrub-volume-data.csv` (note: that above we named this file, `ShrubData`: you many have given it a different name). To learn how to use the `select()` function click on the `select()` link above.

**HAND IN. [Q2]** Add a line of code to your R script where you use the [`mutate()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#adding-a-new-column) function to create a column called `area`. (Reminder, the area of a rectangle is equal to the length * width).

**HAND IN. [Q3]** Add a commented out line of code to your R script that answers the following question: does the `arrange()` function: (a) reorder all the rows in `ShrubData` based on the ascending order of the value in the `width` column, or (b) reorder only the `width` column? (for example, write `# Q3. (a)` in your R script if your answer is (a).)

**HAND IN. [Q4]** Add a line of code to your R script where you use the [`arrange()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#sorting-the-data) function to sort the data based on the `length` column.

**HAND IN. [Q5]** Add a line of code to your R script, where you use the [`filter()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#filtering-data) function to filter the data to only include plants with length longer than 5 cm.

**HAND IN. [Q6]** Add a line of code to your R script, where you use the [`filter()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#filtering-data) function to show all entries with plants with length greater than 4 cm and width greater then 2cm.

**HAND IN. [Q7]** Add a line of code to your R script, where you use the [`filter()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#filtering-data) function to filter the data to only include plants from experiment 1 *or* 2.

**HAND IN. [Q8]** Add a line of code to your R script  where you use the `filter()` function [remove the rows with `NA`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#removing-na-entries-in-data) from your data.


## Mean maximum monthly temperature in Bay D'Espoir

In this section, we want you to use `max-temperature.csv` which you have already downloaded from Brightspace.


> **Metadata Overview** The data are from [Bay D'Espoir GEN STN](https://climate.weather.gc.ca/climate_data/daily_data_e.html?hlyRange=%7C&dlyRange=1967-11-01%7C2019-11-18&mlyRange=1967-01-01%7C2006-02-01&StationID=6567&Prov=NL&urlExtension=_e.html&searchType=stnProv&optLimit=yearRange&StartYear=1840&EndYear=2021&selRowPerPage=25&Line=18&Month=11&Day=30&lstProvince=NL&timeframe=2&Year=2019). and are originally sourced from [Environment and Climate Change Canada](https://climate.weather.gc.ca/historical_data/search_historic_data_e.html). Bay D'Espoir general station is located in Newfoundland at -55.48 W Longitude and 47.59 N latitude. The station Climate ID number is 8400413 . The dataset provided, `max-temperature.csv`, is a cleaned version of the data download.
>
> **Columns in the data frame**
>
> Column 1: Year - The year of the recording
>
> Column 2: Month - The month of the recording, i.e. 1=January; 2=February; etc
>
> Column 3: Day - The day of the recording
>
> Column 4: Max.Temp.C - The maximum temperature recorded on that day in degrees Celcius.


**HAND IN. [Q9]** In your R Script, indicate with a comment that you are starting a new section `## Maximum monthly temperature`. Add lines of code to your R Script where you: 

- [load](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#loading-or-importing-data) the data file: `max_temperature_data.csv`

- remove the `NA`s from the dataset using [`filter()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#removing-na-entries-in-data). 

- calculate the mean maximum temperature in each month across years using `group_by()` and `summarize()` as described in [Group summary](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#group-summary)

- plot the monthly maximum temperature using [`plot()`](https://ahurford.github.io/quant-guide-all-courses/graph.html#scatter-plots-with-plotxy).


```{r, include=F}
temperature.data <- read.csv(file = "lab_data/max-temperature.csv")
monthly_temp <- filter(temperature.data, !is.na(Max.Temp.C)) %>% group_by(Month) %>% summarise(mean = mean(Max.Temp.C))
plot(monthly_temp$Month,monthly_temp$mean)
monthly_temp = as.data.frame(monthly_temp)
```


## Bird Banding

The number of birds banded at a series of sampling sites has been counted by your field crew and entered into the following vector. 

```{r}
number_of_birds <- c(28, 32, 1, 0, 10, 22, 30, 19, 145, 27, 
                     36, 25, 9, 38, 21, 12, 122, 87, 36, 3, 0, 5, 55, 62, 98, 32, 
                     900, 33, 14, 39, 56, 81, 29, 38, 1, 0, 143, 37, 98, 77, 92, 
                     83, 34, 98, 40, 45, 51, 17, 22, 37, 48, 38, 91, 73, 54, 46,
                     102, 273, 600, 10, 11)
```
Counts are entered in order and sites are numbered starting at one.


In your R Script, indicate with a comment that you are starting a new section `## Bird banding`. Review the section [Useful functions](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#useful-functions).
 
 - Add lines of code to your R Script where you provide the commands to answer the questions below.
 
 - In a comment after each line of code for each question, write a comment giving
 the numerical answer, i.e., `# the number of sites is 10.`

**HAND IN. [Q10]** How many sites are there?

**HAND IN. [Q11]** What is the total number of birds counted across all of the sites?

**HAND IN. [Q12]**  What is the smallest number of birds counted?

**HAND IN. [Q13]** What is the largest number of birds counted?

**HAND IN. [Q14]**  What is the mean number of birds seen at a site?

**HAND IN. [Q15]** How many birds were counted at the last site? Have the computer choose the last site automatically in some way, not manually entering its position as a number. Do you know a function that will give you a position of the last value? (since positions start at 1 position of the last value in a vector is the same as its length).

**HAND IN. [Q16]** How many birds were counted at site 42?


```{r, include=F, echo=F}
# Answers to the questions

#how many sites
length(number_of_birds)

#how many birds at site 42
number_of_birds[42]

#what is total number of birds across all sites?
sum(number_of_birds)

#what is smallest number of birds counted?
min(number_of_birds)

#what is largest number of birds counted?
max(number_of_birds)

#what is average number of birds per site?
mean(number_of_birds)

#how many birds were at the last site (without entering position)
number_of_birds[length(number_of_birds)]

```

## Portal Data Manipulation

In this section, we want you to use `surveys.csv`, which you have already downloaded from Brightspace.

```{r, include=F}

survey <- read.csv("lab_data/surveys.csv")

```

In your R Script, indicate with a comment that you are starting a new section `## Portal data`.

**HAND IN. [Q17]** Add lines of code to your R script to [load](https://ahurford.github.io/quant-guide-all-courses/data-entry.html#loading-or-importing-data) the file `surveys.csv`.

**HAND IN. [Q18]** Add lines of code to your R script where you use `filter()` to create a new data frame `survey_new` where you have [removed all rows with `NA`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#removing-na-entries-in-data) in the `weight` column.

**HAND IN. [Q19]**  Add lines of code to your R script where you use the [`select()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#selecting-a-column) function to create a new data frame called `survey_new_2` that only contains the columns `year`, `month`, `day`, and `species_id` in that order.

**HAND IN. [Q20]**   Add lines of code to your R script where you use [`mutate()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#adding-a-new-column) to calculate `weight_kg`, which is weight in **kilograms**, and add this new column to `survey_new_2`, where the result is named `survey_new_3`. The weight in `survey_new_2` was given in grams, so you will need to convert `weight` into kilograms. (*Recall that 1 g = 0.001 kg.*).

**HAND IN. [Q21]**  Add lines of code to your R script, where you use [`filter()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#filtering-data) to make a new data frame `survey_new_4` which is the same as `survery_new_3` except that only the rows with `species_id =="SH"` are included.

**HAND IN. [Q22]** Add a line of code to your R script where you print the first 6 rows of `survey_new_4` using the [`head()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#useful-functions) function.

## Using pipes

In your R Script, indicate with a comment that you are starting a new section `## Using pipes`.

**HAND IN. [Q23]** Add lines of code to your R script were you produce a data frame `survey_new_5` from `surveys.csv`, which is the same as `survey_new_4`, but where you use the [pipe `%>%`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#pipes) operation and do not create the intermediaries `survey_new_2`, `survey_new_3`, and `survey_new_4`.

**HAND IN. [Q24]** Add a line of code to your R script, where you print the first 6 rows of `survey_new_5` using the [`head()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#useful-functions).

```{r, include=F}
survey_new_5 = filter(survey,species_id == "SH") %>%
filter(!is.na(weight)) %>%
            mutate(weight_kg=weight/100) %>%
            select(year, month, day, weight_kg, species_id)
head(survey_new_5)
```


